---
title: "Visualizating Your Data"
description: "Visualizations using ggplot"
toc: true
toc-depth: 3
format:
  html:
    embed-resources: true
    fig-format: png
    fig-dpi: 96

filters:
  - panel-accordion
---
```{r}
#| include: False
library(tidyverse)
```

Please download this [Week 7 Download](downloadforstudent/week_6_studentversion.qmd) to get the .qmd for this weeks lesson. Practice importing this file into your project for this series!

## {tidyverse} packages refresher

The {tidyverse} is a *collection* of R packages that are designed to work together smoothly, so once you learn one, the others feel familiar. 

**Packages within this collection include:**

::: {.panel-accordion}

### dplyr
Houses lots of functions to help with data manipulation. Helps you filter, select, group data variables, and aggregate data. This is a great package for data wrangling.

Cool stuff you can do! 

*Note:* there are a lot of different preloaded datasets to play around with, like mtcars, iris, and (my personal favorite) penguins
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean_mpg = mean(mpg))

```


### ggplot2
Create custom data visualizations, compatible with themes and other custom functions. (this is what we will be focusing on today!)

Cool stuff you can do! 
```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
  geom_boxplot()
```
### tidyr
Data tidying and shaping. Helps you pivot, separate, join, and organize data structures.

Cool stuff you can do! 
```{r}
pivot <- pivot_longer(airquality, 
             cols = c(Ozone, Wind), 
             names_to = "variable", 
             values_to = "value")
head(pivot)
```
### readr
Reads in data files (like CSVs).

Cool stuff you can do! 
```{r}
# Example using a built-in readr dataset
read_csv(readr_example("mtcars.csv"))
```
### tibble
Creates a more modern version of a dataframe that is cleaner and easier to use.

Cool stuff you can do! 
```{r}
tibble(
  x = 1:5,
  y = x^2
)
```
### forcats
Works with **cat**egorical variables. Lets you build more customizable plots.

Cool stuff you can do! 
```{r}
fct_reorder(iris$Species, iris$Sepal.Length) %>%
  head()
```
### stringr
Makes it easy to detect, extract, replace, and clean up strings (text data).

Cool stuff you can do! 
```{r}
str_detect(c("apple", "banana", "pear"), "a")
```
### purrr
Lets you repeat operations without writing loops. Great for applying a function to many things at once.

Cool stuff you can do! 
```{r}
map_dbl(1:5, ~ .x^2)
```
### lubridate
Helps you parse, extract, and do math with dates and times.

Cool stuff you can do! 
```{r}
ymd("2024-02-01") + days(10)
```
:::

## Visualing using {ggplot2}

We will be using datasets from the base package to show off some cool stuff we can do with {tidyverse} packages. Firstly, lets load in the {tidyverse} package. Then, lets use the function `head()` to inspect the preloaded dataset called **state.x77**. 

```{r}
head(state.x77)
```


```{r}
df2 <-cbind(state.x77, data.frame(state.division),data.frame(state.region))

a <-df2 %>%
  group_by(state.region) %>% 
  summarize(IlliteracyRate=mean(Illiteracy, na.rm=TRUE)) #someone walk us through what we're doing here
```

{ggplot2} is a wonderful package in {tidyverse} that creates some very customizable visualizations:

```{r}
#| fig.show: "show"
aplot<- ggplot(a, aes(x=state.region, y=IlliteracyRate, fill=IlliteracyRate)) + #sidenote: why the plus? shouldnt it be %>% ?!
  geom_bar(position=position_dodge(), stat="identity")

aplot
```

```{r}
#| fig.show: "show"
b <-df2 %>%
  group_by(state.division) %>%
  summarize(
    IlliteracyRate=mean(Illiteracy, na.rm=TRUE),
    AvgIncome=mean(Income, na.rm=TRUE)
    )

bplot<- ggplot(b, aes(x=state.division, y=IlliteracyRate, fill=AvgIncome))+
  geom_bar(position=position_dodge(), stat="identity")

bplot
```

```{r}
#| fig.show: "show"
df2plot <- ggplot(df2, aes(x=Income, y=Illiteracy,color=state.region))+ 
  geom_point(stat="identity", size=4, shape=18)

df2plot
```
