---
title: "Getting Started with R"
format: html
description: "Introduction to R, RStudio, and basic syntax"
toc: true
toc-depth: 3
---

Please download this [Week 1 Download](downloadforstudent/week_1_studentversion.qmd) to get the .qmd for this weeks lesson. You will be creating your own project and practice importing this weeks file into your project.

# Getting Situated

Before writing any code, lets get situated with the Rstudio interface. There are **four main panels** in RStudio you will need to become familiar with:

1.  **Script**: this is found in the *top left* of your screen. This is where you write and save your code. This is where you will be working in the most.

    -   To save code, you need to press the file icon at the top of your script, this icon looks like a small disk drive.

2.  **Console**: this is found in the *bottom left* of your screen. This is where R *runs* your code.

    -   You will notice that there are different tabs in the bottom left pane, each of these have different purposes as well. You won’t use these other tabs in Week 1, but it becomes essential for reproducible workflows and GitHub!

        -   **Terminal:** This tab gives you access to your computer’s command line (e.g., Bash, PowerShell). You can use it for tasks like: Navigating folders, Running Git commands, Installing system tools, Managing files outside of R.

        -   **Background Jobs:** This tab shows background tasks that RStudio is running. You’ll see it used when you: knit a large Quarto document, run a long simulation, render a website, run code that RStudio offloads to a separate process. Overall, It’s helpful because it keeps long-running tasks from freezing your main session.

3.  **Environment**: found in the top right of your interface. This shows objects you’ve created. We will unpack what objects are in this lesson.

4.  **Files/Plots/Packages/Help:** Finally, this is found in our bottom right of your screen. Everything in this pane helps you navigate your current project, understand the tools on R, and this is the pane that visualizes any outputs you render.

    -   **Files:** This is one of the most important things to understand how to read as a beginner. Files shows you the files and folders in your current working directory.

        -   **Exercise 1.** type `getwd()` into your Console. What happens?

    -   **Plots:** Whenever you create a plot (e.g. using `plot()` , `ggplot()` , or Quarto visualizations), it will appear here. This is not something you need to worry about now but helpful to know its there.

    -   **Packages:** This tab lists every R package installed on your computer. We will get into what a package is later into this course! This tab (although I don't personally use it very often) shows you see which packages are currently loaded, load or unload packages with a checkbox, install new packages, update existing packages, view package documentation.

        -   *Note:* You can also check these things via commands in your console as well.

    -   **Help:** Whenever you type a function with a ? before it (e.g. `?mean()`) the documentation for that function appears in this pane. This is extremely useful to: read function descriptions, see examples, check argument names, understand error messages, and explore package documentation. You will be using this a lot!

# Let's Make a Project

First, lets go to File -\> new project

There are different ways you can create your working directory folder:

1.  New Directory → creates a brand‑new folder on your computer
2.  Existing Directory → turns an existing folder into an R project
3.  Version Control → if you’re cloning from GitHub

For todays purposes, we will start with a new directory.

Name your project, then choose where you want it to be saved. This is part of your working directory path.

We can create files within your exisiting project, we can also import an existing file. Today we will practice importing this file into your exsiting project. You have a few options:

1.  **Drag and Drop (simplest method)**

-   Open your project folder in Finder/File Explorer

-   Drag your .qmd file into the folder. It will appear in the RStudio Files pane

-   Click it to open

2.  **Use RStudio’s “Upload” or “Import”**

-   In the Files pane click Upload (if using RStudio Server) or Click More → Show Folder in New Window (desktop)
-   Move your .qmd file into the folder
-   Refresh the Files pane
-   Click the file to open it

# Basic R use

R can simply be used as a calculator

```{r}
# Basic math 
5+3
2*6
18/2
4^8

# long operations 
(3+5)*2+5
```

But wait- Those value weren't saved anywhere! To keep them in memory, we will need assign them to a variable.

Let's create a variable called x, that is assigned the number 9. *Note:* A variable is a name that we use to store values. 

```{r}
x <- 8 #this is R syntax 
x = 8 

# What is the value we assign to x? 
x

# What about y? 
#y

# We will need to assign something to y! 
y <- 2 
y

# And we can do things with these variables: 
x+y 
x*y
```

You can use any name for variables, but object names need to start with a letter, and can only contain letters, number, "\_"and "."

## Object and File Formatting:

```{r}
snake_case <- 1
CamelCase <- 1
some.people.like.periods <- 9
Or_STRANE.combiNtions <- 0 # this would be really hard for other people to read your script ;)
```

## Vectors
We can also assign a range of values. Vectors are just 1-dimensional lists of numbers. VECTOR (atomic vector) = a collection of objects of the same class

*Note:* most other programming languages refer to this as an array

```{r}
x <- 1:9 #x is now a vector of numbers 1 through 9 
x
```
 *Also note:* x is no longer number 8, it take on whatever the most assignment

```{r}

# another way to assign a vector 

x <- c(2,3,4,5)
x
```

So far we only use one type of variable (class attribute of an R object), in R it's called class. What classes are those vectors?

## Variable classes (types)

R's basic data types are character, numeric, integer, complex, and logical. You can also embed plots, for example:

```{r}
a <- "12345"
class(a)

b <- 12344
class(b)

c <- c("1", "2", "3")
class(c)

d <- 1:10
class(d)


e <- c(1.1, 2.2, 3.3) +1
class(e)


# Sequence of numbers, 1 thru 10, by 2
seq(from = 1, to = 10, by = 2) #can say "from", "to", and "by=", but they're not necessary
seq(1,10,2)
```

**Exercise 2.** Create a variable called x that is assigned the number 8. Create a variable called y that is a sequence of numbers from 2 to 16, by 2. Add x and y. What happens?

```{r}
x <- 8
y <- seq(2,16,2)
x+y
```

**Exercise 2.a.** Create a variable called z that is a vector containing the numbers 5 and 10. Create a variable called q that is a sequence of numbers from 5 to 25, by 5. Multiply z and q. What happens?

```{r}
z <- c(5,10)

q <- seq(5,30,5)

z*q
```

## Functions

`seq` that we used above is a **function**. Everything you typically want to do in statistical programming uses functions. Functions can take different **arguments** (input).

For example, `sqrt` is a function that can take one numeric argument. We are going to **apply** the function `sqrt` to the value 16.

```{r}
sqrt(16)


# I could have assigned 16 to a variable first:
x <- 16
# And then taken the square root of that variable:
sqrt(x)
```

To help you learn how to use a function you can use `?sqrt` (try it!)

We can also use function to do things for every item in the vector

```{r}
#?sqrt()
# Create a vector:
x <- 1:10
# Get the square root of every number in that vector:
sqrt(x)
```

Another good example is the function `mean`. `mean` takes one **argument**, a numeric vector. We are going to create a vector called `z` and **apply** this function to that vector.

```{r}
z <- 0:20
mean(z)

# Now, let's get the mean of this vector:
q <- c(2, 8, 6, NA, 4, 8)
q
mean(q)
```

Is that the answer you'd expect? let's get some information about the function `mean`

```{r}
#?mean
```

We need to add an additional argument to tell the function that we want to ignore NAs (the default for this argument is FALSE, that's why NAs weren't ignored above).

```{r}
mean(q, na.rm = TRUE)
```

**Exercise 3.** R has a function called `rnorm` that will allow you to get a random sample of numbers drawn from a normal distribtuion. Get a sample of 5 numbers with a mean of 0 and a standard deviation of 0.5.

```{r}

```

There are multiple ways to create arrays of number. As before, we can get a sequence of numbers, 1 thru 10, by 2

```{r}
s <- seq(1,10,2) #can say "by=" before the 2, but it's not necessary
# Sequence of 11 equally spaced numbers between 0 and 1
s2 <- seq(0,1,length.out=11) 
```

*Side note:* what if I don't want these things in my environment any more?

```{r}
rm(x)
rm(q)
rm(s)
# Remove everything
# rm(list=ls())
```
