---
title: "Packages, Data Classes and Indexing"
format: html
description: "Delving into installing and loading packages, data structures, and basics of indexing dataframes"
toc: true
toc-depth: 3
---

Please download this [Week 2 Download](downloadforstudent/week_1_studentversion.qmd) to get the .qmd for this weeks lesson. You will be creating your own project and practice importing this weeks file into your project. (note to self, need to fix link)

## What Is a Package/Library?

Directories are the location of files. Packages/libraries on R are optional set of functions that you can attach to the coding environment. You can think of them as toolkit you can freely download to do complicated and advanced stuff. To use them, you need to first install it. Then, you can load to the current session. Libraries are constantly getting created and updated. Once you pass the basic level, using R(analyzing data) becomes equivalent to "using packages efficiently". So, once you identify your specific issues to solve, always look for packages related to your issue.

**Let's walk through how to install and load a package into your environment:**

**Step 1:** install the package in your console

```{r}
#install.packages("dplyr")
```

The line above downloads the package called dplyr (which we will study together soon!). install.packages("name of package") is the syntax. Occasionally, you may want to set install.packages("whatever",dep=TRUE). "dep" stands for dependency, and it forces R to download all "related" packages to the requested one.

You can also download multiple packages at the same time, with the help of a function we learned before: c()

```{r}
#install.packages(c("tidyverse", "broom"))
```

**Step 2:** Load your library in your script

To use your packages in a given session, you need to load libraries. You can think about installing a package as 'checking a book out' into your own personal library, and when it comes time for you to 'read' it you will need to pull it off the shelf by loading it into your environment.

```{r}
library(dplyr)
library("dplyr")
```

Library should be colored in blue on Rstudio, which indicates this is a special syntax like "for" or "if" or "while". You can provide either character(with "") or just a name.

**Things to be careful when you use packages:** There are few important things to keep in mind when you use libraries.

-   *Make sure that function names are not overlapping.* Each library is often independently developed, so they may contain functions with a same name but different contents. When you load functions with same names from different packages, later ones are prioritized. So, what do we do? To resolve this, you can specify which library is the source of a given function. Say two packages ("packA" and "packB") contain a function called "banana". You can write "packA::banana" to use banana from packA, and vice versa.
-   *Remember that loaded libraries do not disappear even after "rm(list = ls(all = TRUE))".* After you found some form of interference mentioned above, you may want to start over the analysis. To completely start over, you need to shutdown Rstudio because libraries are still loaded even after you clear Environment(workspace)

[Here is a resource](https://support.posit.co/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages) I found online that compiles a list of useful R packages and why they are useful. In general, if you want to do something in R- there is a package for it so always try to look for them first!

**Exercise 1:** Look up the function name `lmer()` using the R help command you learned. What happens? Search the internet and figure out what package the function is from. Install the package and load it.
