---
title: "Let's traverse the TIDYVERSE!"
format: html
description: "Introduction to Tidyverse packages"
toc: true
toc-depth: 3
---
## What is {tidyverse}?

The {tidyverse} is a *collection* of R packages that are designed to work together smoothly, so once you learn one, the others feel familiar. 

## Let's work with {tidyverse}!

Start off with loading in the package: 
```{r}
library(tidyverse)
```

And pick up where we left off last week working with basic dataframes. First, we will define the data.frame:

```{r}
# Define the data.frame
People <-    c("Ulrich","Dom","Sophia","Neha","ExampleRA")
Positions <- c("PI", "GE", "GE",   "GE",  "RA") 
Sleep <-     c(4,6,8,7,8) #Sleep in hours
Years <- c(25,6,2,2,1)#Years in our lab
Count <- c(1:length(People))#should indicate row number

df <- data.frame(People, Positions, Sleep, Years, Count)

head(df)
```

Now, let's do everything we just did with dataframes, but using tidyverse functions

We can use the package {dplyr} to group the data and aggregate by mean. Let's break it down first:

-   **`group_by()` function:** tells R how to organize your data before you do any calculations. We're essentially spliting our data into different groups to look closer at differences between groups.

```{r}
df %>% 
  group_by(Positions)
```

-   **`summarize()` function:** takes each group created by `group_by()` and computes something for it. (can be a mean, median, count, etc.). Where `group_by()` organizes the data but doesn't do any calculation, `summarize()` looks inside each group specified and computes a summary of whats inside.

```{r}
df %>% 
  group_by(Positions) %>% 
  summarise(mean_sleep = mean(Sleep, na.rm = T)) #note: summarize and summarise are the same function, dif spelling 
```

-   **piping, i.e., `%>%`:** The pipe connects steps together so you can read the code like a sentence. It makes your code flow top‑to‑bottom instead of nesting functions inside each other.

We use piping so that we can perform multiple functions on a single dataframe. where your dataframe is always the first argument in tidyverse functions- piping automatically fills that argument for you

Format:

```{r}
#dataframe %>% 
 # some_function() %>% 
 # the_next_function()
```

### How to extract or index variables

Let's using some new functions found in {tidyverse} packages to extract or index variables

```{r}
df$Positions # Get a column "Position"
df[,c("People","Positions")]
cbind(df$People,df$Positions) # Get multiple columns ("People" and "Position")
```

In the {tidyverse}, you can use `select()`

```{r}
df %>% 
  select(Positions) 

df %>% 
  select(People, Positions)
```

**Exercise 1.** Create a new dataframe called 'df_new' that contains only People, Positions, and Sleep (using `select()`)

```{r}

```

### Adding and removing variables (columns)

You can directly add variables

```{r}
df$RowNum <- 1:nrow(df)
```

or directly remove

```{r}
df$RowNum <- NULL
```

But in the {tidyverse}, you can use `mutate()` and `select()` (for adding and removing, respectively)

```{r}
df <- df %>% mutate(RowNum = 1:nrow(.), Years_2026 = Years + 1)

df <- df %>% select(-RowNum)

```

### Filtering rows by condition

Filtering is an operation to keep or remove certain rows

For example, we can filter rows corresponding to GEs

```{r}
df_only_GE <- df[Positions == "GE",]
```

In the {tidyverse}, you can use `filter()`

```{r}
df_only_GE <- df %>% 
  filter(Positions == "GE" | Sleep >= 6)
```

### Why bother with the {tidyverse}?

So far, everything we learned we also learned a very similar way to do using base R functions. So why should we even use {tidyverse}?

{tidyverse} (particularly piping) is helpful for stacking functions:

```{r}
df %>% 
  select(-Count) %>% 
  mutate(sleep_ideal = Sleep + 2) %>% 
  filter(Positions == "RA")
```

It's also helpful for aggregating:

*What?:* Aggregation means taking lots of individual data points and combining them into a smaller set of summary values. *Why?* We do this becuase raw data is often too detailed to interpret directly. It also helps us see patterns more clearly in the data, prepare data for more complicated statistical models (e.g. ANOVA, regression, t-tests), and helps us to communicate results clearly to a wide audience. 

*How?:*

For aggregation, we need to use `group_by()` and `summarize()`

Let's aggregate the average Sleep time for different Positions (PI,GE,RA)

```{r}
df %>%
  group_by(Positions) %>% 
  summarize(meanSleep = mean(Sleep, na.rm = TRUE))

# Sum of Sleep time for different Positions (PI,GTF,RA)

df %>% group_by(Positions) %>% 
  summarize(sumSleep = sum(Sleep, na.rm = TRUE))
```

What happens if you use People as a grouping variable?
```{r}

df %>% group_by(People) %>% 
  summarize(meanSleep = mean(Sleep, na.rm = TRUE)) # Since people is unique for every row, data will not be reduced at all!
```


What if we want to get mean Sleep time for (Group1 = PI & GE, Group2 = RA)

```{r}
#?ifelse()

df2 <- df%>% 
  mutate(PositionsR = ifelse(Positions == "RA", "RA", "PI&GE")) %>%
  group_by(PositionsR) %>% 
  summarize(meanSleep = mean(Sleep, na.rm = TRUE))
```

## More fun with {dplyr}!

We will be using datasets from the base package to show off some cool stuff we can do with {tidyverse} packages.

```{r}
head(state.x77)
```


```{r}
df2 <-cbind(state.x77, data.frame(state.division),data.frame(state.region))

a <-df2 %>%
  group_by(state.region) %>% 
  summarize(IlliteracyRate=mean(Illiteracy, na.rm=TRUE)) #someone walk us through what we're doing here
```

{ggplot2} is a wonderful package in {tidyverse} that creates some very customizable visualizations:

```{r}
#| fig.show: "show"
aplot<- ggplot(a, aes(x=state.region, y=IlliteracyRate, fill=IlliteracyRate)) + #sidenote: why the plus? shouldnt it be %>% ?!
  geom_bar(position=position_dodge(), stat="identity")

aplot
```

```{r}
#| fig.show: "show"
b <-df2 %>%
  group_by(state.division) %>%
  summarize(
    IlliteracyRate=mean(Illiteracy, na.rm=TRUE),
    AvgIncome=mean(Income, na.rm=TRUE)
    )

bplot<- ggplot(b, aes(x=state.division, y=IlliteracyRate, fill=AvgIncome))+
  geom_bar(position=position_dodge(), stat="identity")

bplot
```

```{r}
#| fig.show: "show"
df2plot <- ggplot(df2, aes(x=Income, y=Illiteracy,color=state.region))+ 
  geom_point(stat="identity", size=4, shape=18)

df2plot
```
