---
title: "Let's Work With Dataframes"
format: html
description: "Finding, importing, and working with a dataset"
toc: true
toc-depth: 3
---

Please download this [Week 3 Download](downloadforstudent/week_2_studentversion.qmd) to get the .qmd for this weeks lesson. Practice importing this file into your project for this series!


## Find the data

What is a directory? Directories are the location of files.

**How do I check the current working directory?** For each session (i.e., every time you open Rstudio), the current working directory is set as the default location. You can change this using [this guide](https://stackoverflow.com/questions/22432344/how-do-you-change-the-default-directory-in-rstudio-or-r). When you have an R project open, the working directory is the location of the project. When you have don't have an R project, have R closed and open R by double clicking on a script, then R will open a session with the working directory being the location of the file. If you open R by just clicking on the R icon, it will open a new session with some kind of default location (for windows, this is often "Documents"). 

Let's check our working directory first. 

```{r}
getwd() #print it in a command window 
dirWD <- getwd() #store the current directory as dirWD
```

Specifying a working directory is important for loading and saving files, and it is recommended to define at the beginning of sessions. In R, directories are expressed in the form of string variable like:

```{r}
# This defines a string variable called "dir1".

dir1 <- "C:/Users/forrest/OneDrive/Documents/"
```

*Note:* that each folder is separated by "/". Pressing the tab key may help selecting the right folder/file without typing errors.

How do I change the current working directory? To change the current working directory, you need to specify the directory to jump to. And, feed it in to `setwd()` function (you could also give a string without making a variable).

```{r}
setwd(dir1) #changing directory
getwd() #check if it worked or not
```

**To read in a datafile, you will need to set the working directory to the folder containing the "data" folder:** The path types we have talked about are absolute paths. They are called absolute paths because their origin is at the root location of YOUR system. This means that if you want to share your script #with someone else, it won't work (because they might place files at a different location compared to you). That's why it's important to use R projects or open scripts by opening a fresh R session by double clicking on the script - that way, the working directory location is consistent. So, what if #we need to find a file that's not levels down in our hierarchy, but up?

What's is available in our working directory `fs::dir_tree(path = "..", recurse = TRUE)` If we want to find a file that requires us to go up one level in the hierarchy, we can do the following "../experiment/" or we can set the wd one level higher: setwd("../") should you need to.

## Import The Data
There are many different forms datafiles can take. There are different ways of reading in different files (or different packages that help read in certain datafiles). 

```{r}
#| warning: false
#| message: false
#| results: 'hide'   # hide printed output


#txt files 
txt <- read.table("data/dogwood.txt", header = TRUE, fill = TRUE)

#csv files 
csv <- read.csv("data/2021_demographic.csv")

#spss files 
library(foreign) 
library(tidyverse) 
library(rio)

spss <- as.data.frame(read.spss("data/Experiment_1.sav"))

#RDS files 
rds <- readRDS("data/CCdata.RDS")

#creates results directory 
dir.create("results")

#SAVE FILES 
saveRDS(txt, "results/data2.RDS") 
write.table(txt, "results/data2.txt", sep = "\t") 
write.csv(txt, "results/data2.csv") 
write.foreign(txt, "results/data2.sav", package = "SPSS", codefile = "code.txt")
```

Next, let's read in the .txt file dogwood.txt (data/data2.RDS) and assign it to the name data. 

```{r}
data <- import("data/dogwood.txt")
```

## Working with data

First....let's look at the data! The `$` sign is your new best friend. It let's you index into a dataframe without needing to know the position of the column.

**Exercise 1.** Try typing `data$` to see all of your column names!

```{r}

```


There are lots of ways to index dataframes. You should know them all! 

```{r}
#| warning: false
#| message: false
#| results: 'hide'   # hide printed output


data$ID
data[1]
data$Trial[5] 
data[5,3]
data[1,] #note the comma! 
data$ID[1]
```

We can reference the data inside the brackets too

```{r}
#| warning: false
#| message: false
#| results: 'hide'   # hide printed output

data[data$Trial == 5,]
```

**Exercise 2.** Create a dataframe that contains trials 1 through 5. Then create a dataframe that contains trials 1, 2, and 7

```{r}

```

We can also make new columns using the `$`

```{r}
data$New <- 1:nrow(data)
```

Why would we need to do this? Often useful for cleaning data #So let's clean our data! Sometimes data files contain data that we don't want to analyze for various reasons... 

...like the first trial in a block might be counted as practice 

```{r}
#| warning: false
#| message: false
#| results: 'hide'   # hide printed output



data$Trial > 1 

data[data$Trial > 1,] #index based on this logical argument

dataCleaned <- data[data$Trial > 1,] #make sure to save the result!
```

Functions can also be useful for cleaning. Let's remove RTs that are NaN. (What is Nan?) 

```{r}
dataCleaned <- dataCleaned[!is.na(dataCleaned$RT),]
dataCleaned <- dataCleaned[data$RT > mean(data$RT),]
```

Now let's remove certain trials demo of the AND operator 

```{r}
dataCleaned <- dataCleaned[data$Trial != 5 & data$Trial != 8,]
```

We could only look at cue check trials, demo of the OR operator

```{r}
#dataCleaned2 <- dataCleaned2[dataCleaned$LCueCheck == 1|| dataCleaned$RCueCheck ==1,]
```


**Exercise 3.** remove the first block from our data. We will count it as a practice block. 

```{r}

```

## Analyze Your Data!

*Check your understanding:* What is the mean error?

```{r}
summary(data) #An easy way to get a lot of info 
```

The mean is rarely enough on it's own to tell us much... We want to know how values differ across different experimental conditions. 

*Check your understanding:* What is the error rate for each blockType?

Hopefully you can see that this would be a pain to do by hand... Luckily, there is a better way! Enter, the {dplyr} package. Aggregation in R requires the `group_by()` function, and the `summarize()` function. Pipes are weird... but AWESOME.

{dplyr} is part of the {tidyverse} meta-package, which contains packages designed to help with data wrangling, but most importantly, plotting. Piping feeds in the first argument to a function `data %>% group_by(BlockType)` which is the same as `group_by(data, BlockType)` we need to save this to do anything with it... or we can use pipes

```{r}
errorRates <- data %>%
  group_by(BlockType) %>%
  summarise(meanError = mean(Error))
```


This does the same thing without pipes, just harder to read 

```{r}
errorRates2 <- summarise(group_by(data, BlockType), meanError = mean(Error))
```

Usually we do a 2 step aggregation, which means we do this process twice. Once for each subject individually, then a second time using all subjects data.

```{r}
#| warning: false
#| message: false

errorRates3 <- data %>% 
  group_by(BlockType, Block) %>% summarise(meanError = mean(Error)) %>% #now we do it again
  group_by(BlockType) %>% 
  summarise(Error = mean(meanError), Error_SD = sd(meanError)) #be sure to use the new name of your mean column!

```

Again you could try to do this without piping... it starts to get really hard. 

```{r}
#| warning: false
#| message: false

errorRates3 <- summarise(group_by(summarise(group_by(data, BlockType, Block), meanError = mean(Error)), BlockType), Error = mean(meanError), Error_SD = sd(meanError))
```

But wait, this now contains the entire (raw!) dataset! Let's see tidyverse's version of data cleaning!

```{r}
#| warning: false
#| message: false

data <- data %>%
  filter(Block >= 3 & !is.na(RT) & RT < 10000) %>% 
  group_by(BlockType, Block) %>% summarise(meanError = mean(Error)) %>%
  group_by(BlockType) %>%
  summarise(Error = mean(meanError), Error_SD = sd(meanError))
head(data)
```

Most likely, your data cleaning will be much more complex, so you will have to decide whether and when to save a cleaned version of your dataset, if you want to run multiple different analyses with it (instead of doing both, data cleaning, and analysis in one set of pipes). 
